Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-10-23 manu-draft.qmd
adding study area text

@@

@@

@@

### Models

@@

>>>>

-   Importance of seagrass as foundation species
-   Tampa Bay context
-   Climate change stressors
-   Goals/objectives
    -   Describe

@@

### Study area

Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and one of the largest in the Gulf of Mexico. The watershed covers an 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay. Tampa Bay straddles the temperate and tropical boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer months. The watershed is heavily developed and includes over 3 million people with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023]. The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022]. Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management effort has focused on restoring and maintaining seagrass cover [@sherwood2017]. Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barriers) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20]. Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 295 mi$^2$ [763 km$^2$, @beck2023].

The Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico. Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality conditions primarily from nutrient inputs from wastewater and stormwater [@greening2014]. Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries [@tbep0623]. Notably, Old Tampa Bay has restricted circulation from multiple land bridges that traverse the bay [@sherwood2015; @luther2022]. Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023]. By comparison, water quality conditions in Middle Tampa Bay and Lower Tampa Bay are generally better than the upper two bay segments primarily from more frequent water exchanges with the Gulf of Mexico and lower nutrient loading [@tbep0623]. All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m. Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments prior to wastewater regulation, particularly in Old Tampa Bay and Hillsborough Bay [@greening2014; @johansson15].

### Seagrass monitoring and change

The long-term recovery of seagrass habitats in Tampa Bay is a nationally-recognized success story that demonstrates a successful management paradigm. From 1988 to 2016, seagrasses increased 79% (18,376) to a total of acres ()

@@

### Water quality data

EPC, FIM, Pinellas

### Trend analysis

### Links to seagrass

@@

@@

Alternative nutrient pathways and role of macroalgae, competition for resources, shading, etc.
